---
title: "Introduction to R, Part 2: Data Frames, Reading Data, visualizing data with ggplot2"
author: "Ted Laderas!"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


## Assignment (10 points)

Instructions: Turn your `.nb.html` file in with the following assignment finished. Make sure it is named `part2_LASTNAME.nb.html`, where `LASTNAME` is your last name, of course.


1. Load the `smoke_complete.xlsx` file into R using `read_excel`. You'll need to take a look at it to make sure that your arguments are correct.

```{r}
library(tidyverse)
library(readxl)
smoke_complete <- read_excel("data/smoke_complete.xlsx", sheet=1, na ="NA")

```



2. Use `glimpse()` on `smoke_complete`. How many rows does the file have? How many variables does it have? What are the numeric variables in this dataset?

```{r}
glimpse(smoke_complete)

```

There are 8 numeric variables in this dataset (they are called `dbl`, which is short for double, which is a numeric datatype).

The numeric variables are:

- `age_at_diagnosis`
- `days_to_death`
- `days_to_birth`
- `days_to_last_follow_up`
- `cigarettes_per_day`
- `years_smoked`
- `year_of_birth`
- `year_of_death`



3. Produce a scatter plot of `smoke_complete` using `geom_point()`, mapping these variables to the following aesthetics:


`year_of_birth` to the `x` aesthetic, 
`cigarettes_per_day` to the `y` aesthetic, and
`gender` to the `color` aesthetic.



```{r}
ggplot(smoke_complete, aes(x = year_of_birth,
                           y = cigarettes_per_day,
                           color = gender)) +
geom_point()
```


4. Pick two `numeric` variables, and a character variable in the dataset and produce a scatterplot by mapping them to `x`, `y`, and `color`.  Add a title and change the x and y labels to make sense.

```{r}
ggplot(smoke_complete) + 
  aes(x=age_at_diagnosis/365, y=years_smoked, color = gender) +
  geom_point() +
  xlab("Age at Diagnosis (years)") +
  ylab("Years Smoked") +
  ggtitle("Age versus Years Smoked")

```


