---
title: "Assignment 2"
author: "You"
date: "12/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions for Submitting

Submit your knitted html file to Sakai - if you are having issues, you can submit the `.Rmd` file, put please do your best to knit it. 

Make sure to name your assignment "assignment2-LASTNAME.html" and to fill out the author field with your name. 

Try formatting parts of your text with markdown: https://sph-r-programming.netlify.app/reference/markdown/ - that is, try out some *italics* and **bolding**, or ways to format your text.

## Assignment (10 points)

1.  Load the `data/smoke_complete.xlsx` file into R using `read_excel`. You'll need to take a look at it to make sure that your arguments are correct (1 points).

```{r}
smoke_complete <- read_excel()

```


2.  Use `glimpse()` on `smoke_complete`. How many rows does the file have? How many variables does it have? What are the numeric variables in this dataset? (2 points)

```{r}
glimpse(smoke_complete)

```

[Put your answer here]


3. Use `skim()` on `smoke_complete()`.  Is the `year_of_death` variable skewed or not? Describe how it is skewed. (2 points)

```{r}
library(skim)


```

3.  Produce a scatter plot of `smoke_complete` using `geom_point()`, mapping these variables to the following aesthetics:

`year_of_birth` to the `x` aesthetic, `cigarettes_per_day` to the `y` aesthetic, and `gender` to the `color` aesthetic.

(2 points)

```{r}
ggplot(smoke_complete, aes(x = ____,
                           y = ____,
                           color = ___)) +
    ----------
```


4.  Pick two `numeric` variables, and a character variable in the `smoke_complete` dataset and produce a scatterplot by mapping them to `x`, `y`, and `color`. Add a title and change the x and y labels to make sense. (2 points)

```{r}


```

5. There is a dataset called `penguins` in a package called `palmerpenguins` (it has already been installed in your project.) Use `skim()` to skim the dataset, and pick two numeric variables to plot as a scatterplot, and pick one of the categorical or (also known as `factor`) variables to color your plot by. (3 points)

You can find out more about the dataset by using either `?penguins` or `help(penguins)`:

```{r}
#load the palmerpenguins library
library(palmerpenguins)
#load the dataset using data()
data(penguins)
penguins
```

1 Bonus point for an interpretation of your plot.