---
title: "Assignment 2"
author: "You"
date: "12/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions for Submitting

Submit your knitted html file to Sakai - if you are having issues, you can submit the `.Rmd` file, put please do your best to knit it. 

Make sure to name your assignment "assignment2-LASTNAME.html" and to fill out the author field with your name. 


## Assignment (10 points)

1.  Load the `data/smoke_complete.xlsx` file into R using `read_excel`. You'll need to take a look at it to make sure that your arguments are correct.

```{r}
smoke_complete <- read_excel()

```



2.  Use `glimpse()` on `smoke_complete`. How many rows does the file have? How many variables does it have? What are the numeric variables in this dataset?

```{r}
glimpse(smoke_complete)

```

[Put your answer here]



3.  Produce a scatter plot of `smoke_complete` using `geom_point()`, mapping these variables to the following aesthetics:

`year_of_birth` to the `x` aesthetic, `cigarettes_per_day` to the `y` aesthetic, and `gender` to the `color` aesthetic.

```{r}
ggplot(smoke_complete, aes(x = ____,
                           y = ____,
                           color = ___)) +
    ----------
```


4.  Pick two `numeric` variables, and a character variable in the `smoke_complete` dataset and produce a scatterplot by mapping them to `x`, `y`, and `color`. Add a title and change the x and y labels to make sense.

```{r}


```

