---
title: "Assignment 6"
author: "Ted Laderas"
date: "2/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(purrr)
library(readxl)

load_files <- function(path){
  out_frame <- readxl::read_excel(path, na="NA")
  out_frame <- janitor::clean_names(out_frame) %>%
    mutate(source=path)
  return(out_frame)
}

```

# Assignment 6

## Question 1 (2 Points)

Load the following list using `readRDS()` - an RDS file is a way to save a single R object as a file (you can write them using `saveRDS()`). 

What are the names of slots on the list? (show the function)

```{r}
dfs <- readRDS("data/list_structure.rds")


```

What is in each slot? Just put your answer below.

## Question 2 (2 Points)

Use `bind_rows()` on the first and third element of `dfs`:

```{r}
bind_df <- bind_rows(------, -------)
```

## Question 3 (2 Points)

Find the maximum `age_at_diagnosis` for `dfs[["smoke_2.xlsx"]]`: (remember, you can use `summarize`)

```{r}


```

## Question 4 (2 Points)

Write a function that finds the maximum `age_at_diagnosis` on a data.frame. Use it on the second element of `dfs`.

```{r}
find_max_age <- function(df){
  max_age <- df %>%
    ---------
  return()
}

find_max_age()
```

## Question 5 (2 Points)

5. Try out a different theme/options for this `html_document`, or make it into a `distill::distill_article`. Add a table of contents and section numbering.