---
title: 'Part 6: Doing Stats, Making Friends'
author: "you"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Final Project (Due Finals Week)

This is your opportunity to show your knowledge so far.


1. Find a dataset that you're interested in. It should be either in excel or csv formats.
2. Decide on a question you're interested in.
3. Explore the dataset and transform the variables
4. Visualize the dataset and talk about your insights.

Please set up a time to talk with Aaron or me about your data and your goals.


# Learning Objectives

- **Learn** about errors and warnings and where to ask for help
- **Learn** some basic Exploratory Data Analysis techniques
- **Learn** a basic analysis workflow for statistical modeling
- **Learn** about *formulas* and how to specify models using them
- **Learn** about t Tests and how to **apply** them to your dataset
- **Learn** and **apply** linear regression models
- **Learn** and **apply** Analysis of Variance (ANOVA)


# Getting Help on Errors

## Understanding the difference between warnings and errors

A *warning* is an indication that the data or arguments isn't quite what the function expected. 

You can usually run the code, but you should be careful about it and verify the output.

An *error* means that the code can't execute at all given what you have given the function.

Errors can be difficult to understand, which is why



## Googling is StandaRd pRactice foR eRrors

The first thing I do when I encounter an error is to search for the error. I usually start with Google.

I don't know everything, and the odds are that I made a mistake in understanding the documentation.

There are some resources that I especially check (in order):

- RStudio Community (for `tidyverse`): https://community.rstudio.com/
- Stack Overflow: http://stackoverflow.com/
- Biostars (for Bioinformatics): https://www.biostars.org/
- The package's github page (especially issues)



## Where do I ask for help?

I'm trying to be as helpful as I can, but I can't answer all of your questions.

The following communities are extremely helpful to beginners:

- R for Data Science Community: https://r4ds.slack.com/
- RStudio Community: https://community.rstudio.com/




# Caveat about statistics

This is not meant to be a comprehensive course in statistics. We want to show you some basic techniques, but you will need to dig further.

Danielle Navarro's Learning Statistics with R is excellent and talks much more about statistics: https://learningstatisticswithr.com/




# Introducing `tidymodels`

We will be using the `broom` package from the `tidymodels` set of packages to make the modeling easier to work with. 

`tidymodels` attempts to unify all of the various modeling packages in a consistent interface. 

`broom` works mostly with the output of models. One of the problems with R is that the many modeling packages are not consistent to work with. It can be just as difficult to get a p-value out of a model as it is to run it on some data! `broom` simplifies this a lot.

There are 3 main functions in *broom*:

- `tidy()` - This is where you get most of the output you want, including *coefficients* and *p-values*
- `glance()` - additional measures on your model, including R-squared, log likelihood, and AIC/BIC
- `augment()` - make predictions with your model using new data


We will mostly use `tidy()` and `glance()` for right now.



# T-tests

## The Dataset


A study by Goran et.al (1996) examined the accuracy of some widely used body-composition techniques for children using three different methods: 

  - dual-energy X-ray absorptiometry (`dxa`) technique, 
  - skin-fold thickness (`st`), 
  - bioelectric resistance (`br`). 
  
Subjects were children between 4 and 10 years old. Data were collected on 98 subjects (49 males and 49 females). 

One purpose of the study was to determine whether there was a difference in fat mass measurements using `DXA` (considered the gold standard method) compared to the skin-fold thickness method. 

We also wish to determine if `DXA` levels are significantly different between males and females.

## Getting set up

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(readxl)
library(janitor)
library(GGally)
library(broom)
```


```{r message=FALSE}
body_comp <- read_csv('data/body_composition.csv', na="NA") %>%
  clean_names() %>%
  mutate(gender = factor(gender, levels=c("1", "0")))

head(body_comp)
```



## Exploratory Data Analysis

Before we do any statistical tests on our data, we should first visualize it. 

Since our ultimate goal is to examine the differences between bodyfat measurement methods, let's create boxplots that illustrate this difference, if any. 

Notice that the `aes()` for `ggplot()` only accepts one `x` value and one `y` value, but we have three columns we'd like to compare (`dxa`, `st`, `br`). So, we  need to convert our data to long format using `pivot_longer()`.


```{r}
body_comp_long <- body_comp %>%
  pivot_longer(cols = c('dxa', 'st', 'br'),
               names_to = 'method',
               values_to = 'body_fat_percentage') 


head(body_comp_long)
```


Now that we've done that, we can set `x = method` and `y = body_fat_percentage`.

```{r warning=FALSE}
ggplot(body_comp_long) +
  aes(x = method, y = body_fat_percentage, fill = method) +
  geom_boxplot() +
  geom_jitter(color="grey")
```

It appears that our measurements are close to one another, but there are some noticeable differences.



## t-Test

Briefly, a t-Test should be used when examining whether the mean **between two groups** are similar This means that the measurements must be **numeric** (there are other tests for categorical data). 

The null hypothesis for a t-test is that the two means are equal, and the alternative is that they are not.


> One purpose of the study was to determine whether there was a difference in fat mass measurements using `dxa` (considered the gold standard method) compared to the skin-fold thickness method (`st`).


Below, we will use a paired t-test. Paired simply means that each group (`dxa` and `st`) each contain measurements for the same subject on corresponding rows. If body fat measurements were collected using `dxa` for children in Group A and `st` for a separate set of children in Group B, then we would *not* use a paired t-test.


### HYPOTHESIS: There is a difference in mean fat mass measurements between the DXA and skin-fold thickness (ST) methods.

### NULL HYPOTHESIS: There is no difference in mean fat mass measurements between the two methods.


We also need to set a significance threshold. We'll set it at 0.05.

Like many statistical modeling methods in R, the `t.test` function takes the model in a couple different ways. The first takes two separate arguments with the paired measurements: measure 1 (dxa) and measure 2 (st). We use `tidy()` from `broom` to clean up the output into a data frame:

```{r}
tidy_output <- t.test(body_comp$dxa, body_comp$st, paired=TRUE) %>%
  tidy()

tidy_output
```

The second method uses the formula method, where the outcome and grouping variable (dxa vs. st) are specified using special syntax that uses the tilde (`~`) symbol, with the outcome on the left and the grouping on the right of the `~`. The tilde (`~`) can be translated to "is a function of".

Note we need to use the long data filtered to only include the two methods of interest (our grouping variable), and then specify the data frame so `t.test` knows where the variables are coming from:

```{r}
body_comp_dxa_st <- body_comp_long %>%
  filter(method %in% c("dxa", "st"))

tidy_output2 <- 
  t.test(body_fat_percentage ~ method, 
        paired=TRUE, 
        data=body_comp_dxa_st) %>%
    tidy()

tidy_output2
```

We see that `p.value` is approximately equal to `r round(tidy_output2$p.value,3)`; this means **we cannot reject the null hypothesis** (i.e., the difference in body fat measurements between `dxa` and `st` are not statistically different from one another).



## Your Turn

Try running `t.test`, comparing `dxa` and `br` using `body_comp_long`. Hint: You'll have to filter the `method` like above.

```{r eval=FALSE}
body_comp <- body_comp %>%
  tidyr::drop_na()

body_comp_long <- body_comp %>%
  pivot_longer(cols = c('dxa', 'st', 'br'),
               names_to = 'method',
               values_to = 'body_fat_percentage') 


body_comp_dxa_sf <- body_comp_long %>%
  filter(method %in% c("dxa", "br"))

tidy_output <- 
  t.test(body_fat_percentage ~ method, 
       paired=TRUE, 
       data=body_comp_dxa_sf) %>%
    tidy()

tidy_output
```






# How Correlated are the Three Variables?

Another question we'd like to check is whether the measurements are correlated or not. 

That is, can we reliably predict `dxa` from `st`? 

Let's generate a pairs plot, which can be a useful way of visualizing correlated variables. 

We can do this by using `ggpairs()`, which is in the `GGally` package (not part of the `tidyverse`):

```{r message=FALSE}
GGally::ggpairs(body_comp,  aes(color=gender))  
```


We can generate a scatterplot with a trend line using a geom called `geom_smooth()`. We use `method = "lm"` to specify that we want a best fit straight line from the linear model method, not the default loess curve.

We need to add the arguments `method` and `se` (short for standard error).

```{r}
body_comp %>%
  ggplot() +
  aes(x=dxa, y=st) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

```


## Your Turn

Try setting `se` to `TRUE`. What does it add to the graph?

```{r}
body_comp %>%
  ggplot() +
  aes(x=dxa, y=st) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE)
```



## Let's build a simple linear model

We've established that there is a correlation between `dxa` and `st`. Can we use `st` to predict `dxa`?

y = ax + b

The function `lm()` uses the formula method, with the outcome (dependent variable) on the left and the predictors (independent variables) on the right, separated by the plus sign (`+`) if there are multiple predictors. We again specify the data frame with `data = body_comp` so the function knows where the variables in the formula are coming from.


```{r}
lm(dxa ~ st, data = body_comp) %>%
  tidy()
```

The predicted line through the data is:

dxa  = 0.275    +    0.903 * `st`



## Adding another variable


What if we included `gender` in our model? Our model can accept `factors` as inputs.

```{r}
lm(dxa ~ st + gender, data = body_comp) %>%
  tidy()
```

In this case, `gender` is a useful predictor of `dxa`, since the p-value is less than our threshold of 0.05.

However, if we did use the equation, it would correspond to this equation:


`dxa` = 0.097 +   0.889 * `st`    +    0.536 * `gender0`


The factor `gender` here is recoded as a "dummy" variable, and reading it is a little confusing. Note that it says `gender0` and not `gender`. That's because it's coding `gender` = 0 as 1 here, and 0, if gender is 1.

Dummy variables are very confusing. http://www.sthda.com/english/articles/40-regression-analysis/163-regression-with-categorical-variables-dummy-coding-essentials-in-r/

We can make this dummy variable a bit more clear by taking care with the ordering of our factor levels. The reference variable is the first factor level. Right now, `gender` is a factor with levels:

```{r}
levels(body_comp$gender)
```

But we can change the reference group to 0 so that the coefficient is calculated for gender = 1:

```{r}
body_comp <- body_comp %>%
  mutate(gender = fct_relevel(gender, "0"))

lm(dxa ~ st + gender, data=body_comp) %>%
  tidy()
```

Why did the coefficient of gender change signs?

## Your Turn

Try adding `br` as a term in the model. How does it change the p-value of `st`?


```{r}
lm(dxa ~ st + br, data=body_comp) %>%
  tidy()
```




# Analysis of Variance (ANOVA) (Optional)

We've determined that there isn't a statistical difference between `dxa` and `st` using a paired t-test, but we also measured bodyfat using bioelectric resistance, `br`. 

Maybe we should see if it measures differently from the other two methods. Because a t-test can only be used to measure the differences in means between two groups, we'd have to use multiple t-tests.

But wait, should we do that right away? No, because we'll run into the [Multiple Comparisons Problem](https://en.wikipedia.org/wiki/Multiple_comparisons_problem)!

So rather than performing multiple t-tests, we first want to examine whether any of the groups is different from the rest of the groups using an ANOVA (`aov()`).

`aov()` uses the formula interface where the group variable is on the right of the `~`, and therefore we need the long version of our data with `method` denoting the three groups. Below, we are testing whether body fat percentage is a function of the type of body fat measurement method. We pipe the output of `aov()` to `tidy()` to get a clearer idea of the output of the ANOVA.



```{r}
aov(body_fat_percentage ~ method, data = body_comp_long) %>%
  tidy()
```

The value `p.value` is what we're interested in. Because it is greater than 0.05, we can conclude that none of the measurement methods is significantly different from the others, and there is no reason to perform multiple t-tests on our dataset.


## Post-hoc Tests

Since our F statistic probability was not significant at below 0.05, we would not expect post-hoc pairwise t-tests comparing each pair of methods to be significant. But we can still perform multiple post-hoc t-tests using the function `pairwise.t.test()`. However, we need to account for inflation of false positives by using a multiple testing correction method (e.g., Bonferroni).

```{r}
pairwise.t.test(body_comp_long$body_fat_percentage, 
                body_comp_long$method, 
                p.adjust = "bonferroni") %>% 
  tidy()
```


## More about the Multiple Testing Problem

Consider what a p-value of 0.05 actually means: if a test is performed at the 0.05 level and the corresponding null hypothesis is true, there is only a 5% chance of incorrectly rejecting the null hypothesis. This is an okay risk to take given that we are only performing the t-test once. But if we were to perform the t-test 1,000 times on data where all null hypotheses were true, the expected number of incorrect rejections (also known as false positives or Type I errors) would be 50!


# Tidy Models with `parsnip`

We used the `broom` package in the `tidymodels` suite of packages, but there are additional tools to fit models including those in the `parsnip` package. This package attempts to standardize the input of the model functions so that we do not have to remember which functions take formulas such as `y ~ x` and which take separate arguments for `x` and `y`.

We first specify a "specification" of the model before we fit, and we need to specify the mode ("regression" vs "classification" etc) and the computational engine ("lm" vs other functions for fitting models using R, or using other engines such Stan, Spark, or keras, which become useful with big data):

```{r}
library(parsnip)

lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_spec
```

Now we fit the model using `fit()` and `tidy()` it up:

```{r}
lm_spec %>%
  fit(dxa ~ st + br + gender, data = body_comp) %>%
  tidy()
```

We see the same result we would obtain from using `lm()` above. The usefulness of `parsnip` and `tidymodels` really shines when fitting multiple different types of machine learning models (i.e Random Forest, elastic net, neural net), or adding in other regression and machine learning methods such as bootstrap resampling, prediction accuracy estimation, cross-validation, and so on. We recommend studying the vignettes and articles in https://www.tidymodels.org/learn/.

Here's an example of fitting a random forest model:

```{r}
rand_forest() %>%
  set_engine("randomForest") %>%
  set_mode("regression") %>%
  fit(dxa ~ st + br + gender, data = body_comp)
```




## Acknowledgements

Written by Aaron Coyner and Ted Laderas and Jessica Minnier


