---
title: "Introduction to RMarkdown"
author: "Ted Laderas"
date: "2/8/2021"
output: 
  html_document:
    df_print: paged
#    toc: true
#    toc_float: true
#    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
data(penguins)
```

## The HTML document format

The `html_document()` format is extremely powerful and customizable.


## Table of Contents

Try uncommenting the following code and reknit the document.

```
#    toc: true
#    toc_float: true
```

## Themes

Try out the following themes in the `theme:` front matter to change the appearance of the document.

- cosmo
- lumen
- paper

## A Pretty Table

Try changing the argument for `df_print:` to the following formats:

- paged
- kable

```{r}
penguins
```


## {gt} package

There is also the `{gt}` package (**g**rammar of **t**ables), which lets you highly customize tables. I'm not an expert at it, but you can group columns and such.

```{r}
library(gt)
gt(penguins[1:10,])
```

## {gtsummary}

For making table 1s of most papers (the cohort table), `{gtsummary}` is extremely helpful:

```{r}
library(gtsummary)

penguins %>%
tbl_summary(by=species) %>% 
  add_n() %>% # add column with total number of non-missing observations
  add_p() %>% # test for a difference between groups
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels()
```


## Example of a Tabset {.tabset}

### Showing Graph

```{r echo=FALSE}
ggplot(penguins) +
  aes(x = species, y= flipper_length_mm, fill = species) +
  geom_boxplot() +
  facet_wrap(vars(island))

```

### Showing Code

```{r eval=FALSE, echo=TRUE}
ggplot(penguins) +
  aes(x = species, y= flipper_length_mm, fill = species) +
  geom_boxplot() +
  facet_wrap(vars(island))

```

