---
title: "Introduction to Distill"
author: 
  - first_name: "Ted"
    last_name: "Laderas"
    url: https://laderast.github.io
    affiliation: Division of Bioinformatics and Computational Biomedicine, Medical Informatics and Clinical Epidemiology; Knight Cancer Institute, Oregon Health & Science University
    orcid_id: 0000-0002-6207-7068
  - first_name: "Joan"
    last_name: "Rivers"
    affiliation: Multiple Comedy Clubs
date: "2/8/2021"
output: 
  distill::distill_article:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
#note we're setting echo to FALSE here (don't show code)
knitr::opts_chunk$set(echo = FALSE)
library(palmerpenguins)
data(penguins)
library(tidyverse)
library(rmarkdown)
library(plotly)
```

# Welcome to Distill

Distill is a display format meant for scientific papers and sharing on the web. It has a lot of built in features that make it ideal for sharing, including:

- Auto Numbered Tables 
- Auto Numbered figures
- Footnotes
- Appendices
- References

## Distill articles are single files

Much like `html_document` files, `distill` files are easily shared, either by emailing, Box, or on the web with Netlify Drop.

## Table Example

As can be seen in Table \@ref(tab:table1), we have a large number of penguins.

```{r table1}
knitr::kable(penguins[1:10,], caption = "Penguin Table")
```

## Figure Example

As can be seen in Figure \@ref(fig:figure1)), the penguins are not all distributed evenly across the islands. ^[There are a lot of penguins everywhere.]

```{r figure1, fig.cap="Palmer Penguins"}
ggplot(penguins) +
  aes(x = species, y= flipper_length_mm, fill = species) +
  geom_boxplot() +
  facet_wrap(vars(island))
```

## Interactive Figures

The `{plotly}` package lets you make interactive versions of these figures. You can email this entire html document to someone, or you can put it online.

```{r figure3, fig.cap="Palmer Penguins Interactive"}
my_plot <- ggplot(penguins) +
  aes(x = species, y= flipper_length_mm, color = species) +
  geom_jitter() +
  facet_wrap(vars(island))

ggplotly(my_plot)
```

## {gtsummary}

For making table 1 of most papers (the cohort table), `{gtsummary}` is extremely helpful for a 

```{r table2, fig.cap="Table 1 summary of Penguins"}
library(gtsummary)

penguins %>%
tbl_summary(by=species) %>% 
  add_n() %>% # add column with total number of non-missing observations
  add_p() %>% # test for a difference between groups
  bold_labels() 
```


## For More Information about Distill

https://rstudio.github.io/distill

## Appendix Example {.appendix}

Here's an Example of an Appendix.