---
title: 'Introduction to R, Class 4: Data Visualization'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

- create and modify scatterplots and boxplots
- represent time series data as line plots
- split figures into multiple panels using `facet_wrap()`
- customize your plots using built in `theme()`s

- Sort by a variable in a dataset using `arrange()`
- Select variables in a dataset using `select()`
- Filter a dataset using the `filter()` function

## Getting set up

```{r prep}
# load library
library(tidyverse)
library(readxl)
library(here)
# read in second filtered data from last class
smoke_complete <- read_excel("data/smoke_complete.xlsx", sheet=1, na="NA")

smoke_complete <- smoke_complete %>% select(age_at_diagnosis, tumor_stage, cigarettes_per_day, gender, vital_status, disease)
```

## Customizing a Scatterplot

Now that we have the data generally displayed the way we'd like, we can start to customize a plot.

```{r}
our_plot <- 
  
  ggplot(smoke_complete) +

  aes(x = age_at_diagnosis, 
      y = cigarettes_per_day, 
      color = disease) +
  
  geom_point() +
  
  labs(title = "Cigarettes per Day versus Age at Diagnosis",
       x = "Age at Diagnosis",
       y = "Cigarettes Smoked per Day")
  
our_plot
```

# Changing visual properties using built in themes

```{r}
our_plot +
  theme_bw()
```

### Make the points more transparent

```{r theme}
ggplot(smoke_complete) +
  aes(x = age_at_diagnosis, 
      y = cigarettes_per_day, 
      color = disease) +

  geom_point(alpha=0.3) +
  
  theme_bw() # change background theme
```


```{r color}
# change color of points
ggplot(smoke_complete) +
  aes(x = age_at_diagnosis, 
      y = cigarettes_per_day) +
  geom_point(color = "green")
```

For more information on colors available, look [here](http://sape.inf.usi.ch/quick-reference/ggplot2/colour).

A complete list of pre-set themes is available [here](https://ggplot2.tidyverse.org/reference/ggtheme.html), 
and we'll cover ways to customize our own themes later in this lesson.


```{r alpha}
# add transparency with alpha argument
ggplot(smoke_complete) +
  aes(x = age_at_diagnosis, y = cigarettes_per_day) +
  geom_point(alpha = 0.1)
```

Transparency is useful to help see the distribution of data, 
especially when points are overlapping.

Alternatively, you can alter only one specific type of text:
```{r x_text}
ggplot(smoke_complete) +
  geom_point(aes(x = age_at_diagnosis, y = cigarettes_per_day, color = disease)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) # rotate and adjust x axis text
```
This rotates and adjusts the horizontal and vertical arrangement of the labels on only the x axis.
Of course, you can also modify other text (y axis, axis labels, legend).

After you're satisfied with a plot, 
it's likely you'd want to share it with other people or include in a manuscript or report.
```{r dir, eval=FALSE}
# create directory for output
dir.create("figures")
```
```{r save}
# save plot to file
ggsave("figures/awesomePlot.jpg", width = 10, height = 10, dpi = 300)
```

This automatically saves the last plot for which code was executed. 
You can view your `figures/` directory to see the exported jpeg file.
This command interprets the file format for export using the file suffix you specify. The other arguments dictate the size (`width` and `height`) and resolution (`dpi`).

**Challenge:**
Create a scatterplot showing age at diagnosis vs years smoked with points colored by gender and appropriate axis labels 
(solutions [here](solutions/class4_solutions.R))



## Brief Aside: Categorical Data (`factor`s)


One data type that we haven't yet looked at are `factor`s - factors are how R represents *categorical data*.

For the most part, you can use `character` and `factors` interchangably for categorical data.

However, there is one main difference. `factor`s define the permissible values in a vector with an argument called `levels`. They also define the order in which these values are displayed.


```{r}
character_vector <- c("Dog", "Dog", "Cat", "Mouse")
table(character_vector)
```

You can make a character vector into a factor vector by using the `factor()` function, and supplying an argument called `levels`.

## Levels of a Factor

The `levels` of a factor are the permissible values in a `factor`. The order of the values in a factor also control the order in which the values appear in tables and on the axes in a plot. 

So, you can control the order of the categories in a factor by specifying the order of the categories in the `levels` argument.

```{r}
factor_vector <- factor(character_vector, 
                        levels = c("Dog", "Cat", "Mouse"))

table(factor_vector)
```

This ordering is the main reason to use `factors`. We will revisit this again in Part 4 when we learn how to manipulate the data type of a variable in a `data.frame`.




## Boxplots

Boxplots compare the distribution of a quantitative variable among categories.

Remember, `vital_status` is a `character` vector, but we're not too worried about the implicit order of the categories, so we can use it as is in our boxplot.

```{r box}
# creating a boxplot

ggplot(smoke_complete) +
  
  aes(x = vital_status, y = cigarettes_per_day) +
  
  geom_boxplot()
```

The main differences from the scatterplots we created earlier are the `geom` type and the variables plotted.

We can change the color similarly to scatterplots. However, we map to `fill` and not `color`

```{r box_color}
# adding color
ggplot(smoke_complete) +
  
  aes(x = vital_status, 
      y = cigarettes_per_day, 
      fill = vital_status) +
  
  geom_boxplot()
```

## Faceting our boxplot

One of the most powerful ways to change a visualization is by *faceting*: 


```{r box_color}
# adding color
ggplot(smoke_complete) +
  
  aes(x = vital_status, 
      y = cigarettes_per_day, 
      fill = vital_status) +
  
  geom_boxplot() +
  
  ylim(c(0,20)) +
  
  facet_wrap(vars(disease))
```

**Challenge:** 
Write code comments for each of the following lines of code. 
What is the advantage of writing code like this? 
(solutions [here](solutions/class4_solutions.R))
```{r challenge}
my_plot <- ggplot(smoke_complete, aes(x = vital_status, y = cigarettes_per_day)) 
my_plot +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.2, color = "purple")
```

**Challenge:** 
Does the order of layers in the last plot matter? What happens if `jitter` is coded before `boxplot`? 
(solutions [here](solutions/class4_solutions.R))

## Time series data as line plots

So far we've been able to work with the data as it appears in our filtered dataset.
Now that we're moving on to time series plots (changes in variables over time),
we need to manipulate the data.
We'll also be working with the `birth_reduced` dataset,
which we created last class (primarily by removing all missing data for year of birth).
We'd like to plot the number of individuals in the dataset born by year,
so we need to first count our observations based on both disease and year of birth:
```{r time_prep}
# count number of observations for each disease by year of birth
yearly_counts <- birth_reduced %>%
  count(year_of_birth, disease) 
```

We can plot these data as a single line:
```{r time_plot}
# plot all counts by year
ggplot(yearly_counts) +
  geom_line(aes(x = year_of_birth, y = n))
```

Here, `n` represents the number of patients born in each year,
from the count table created above.
The result isn't very satisfying, because we also grouped by disease.
We can improve this by plotting each disease on a separate line, 
which is more appropriate when there are multiple data points per year:
```{r time_disease}
# plot one line per cancer type
ggplot(yearly_counts) +
  geom_line(aes(x = year_of_birth, y = n, 
                group = disease))
```

Moreover, we can color each line individually:
```{r time_color}
# color each line per cancer type
ggplot(yearly_counts) +
  geom_line(aes(x = year_of_birth, y = n, color = disease))
```

Note that you don't have to include a separate argument for `group = disease` because grouping is assumed by `color = disease`.

**Challenge:** 
Create a line plot for year of birth and number of patients with lines representing each gender.
Hint: you'll need to manipulate the `birth_reduced` dataset first. 
(solutions [here](solutions/class4_solutions.R))

**Challenge:** 
How do you show differences in lines using dashes/dots instead of color?
(solutions [here](solutions/class4_solutions.R))

## Faceting 

So far we've been working on building single plots,
which can show us two main variables (for the x and y axes)
and additional variables using color (and potentially size/shape/etc).
Scientific visualizations often need to compare among categories 
(e.g., control vs various treatments),
which is generally clearer if those categories are presented in separate panels.
ggplot provides this capacity through faceting.

Let's revisit the scatterplot we initially created,
plotting age at diagnosis by cigarettes per day,
with points colored by disease.
We add an additional layer to create facets, or separate panels,
for a given variable 
(in this case, the same variable being used to color points):
```{r facet}
# use previous scatterplot, but separate panels by disease
ggplot(smoke_complete) +
  geom_point(aes(x = age_at_diagnosis, y = cigarettes_per_day, color = disease)) +
  facet_wrap(vars(disease)) # wraps panels to make a square/rectangular plot
```

`vars` is used for faceting in the same way that `aes()` is used for mapping:
it is used to specify the variable to form facet groups.

`facet_wrap` determines how many rows and columns of panels are needed to create the most square-shaped final plot possible.
This becomes useful when there are many more categories:
```{r facet_wrap}
# add a variable by leaving color but changing panels to other categorical data
ggplot(smoke_complete) +
  geom_point(aes(x = age_at_diagnosis, y = cigarettes_per_day, color = disease)) +
  facet_wrap(vars(tumor_stage))
```

In this case, we're now visualizing an additional variable (tumor stage), 
in addition to the original three (age at diagnosis, cigarettes per day, and disease).

If you want to control the specific layout of panels, you can use `facet_grid` instead of `facet_wrap`:
```{r facet_grid}
# scatterplots with panels for vital status in one row
ggplot(smoke_complete) +
  geom_point(aes(x = age_at_diagnosis, y = cigarettes_per_day, color = disease)) +
  facet_grid(rows = vars(vital_status)) 
```

This method can also plot panels in columns.

We may want to show interactions between two categorical variables,
by arranging panels into rows according to one variable and columns according to another:
```{r facet_both}
# add another variable using faceting
ggplot(smoke_complete) +
  geom_point(aes(x = age_at_diagnosis, y = cigarettes_per_day, color = disease)) +
  facet_grid(rows = vars(vital_status), cols = vars(disease)) # arrange plots via variables in rows, columns
```

Don't forget to look at the help documentation (e.g., `?facet_grid`) to learn more about additional ways to customize your plots!

**Challenge**: 
Alter your last challenge plot of (birth year by number of patients) to show each gender in separate panels.
(solutions [here](solutions/class4_solutions.R))

**Challenge**: How do you change axis formatting, like tick marks and lines?
Hint: You may want to use Google!
(solutions [here](solutions/class4_solutions.R))

## Wrapping up

This material introduced you to ggplot as a tool for data visualization, 
allowing you to now create publication-quality images using R code.
Combined with our previous explorations of the basic principles of R syntax, 
importing and extracting data with base R,
and manipulating data using tidyverse, 
you should be equipped to continue learning about R on your own and developing code to meet your research needs. 

If you are interested in learning more about ggplot:
- Documentation for all `ggplot` features is available [here](https://ggplot2.tidyverse.org).
- RStudio also publishes a [ggplot cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf) that is really handy!





# Data Manipulation

What we have learned is pretty great overall for plotting. But there are a lot of other things we can do with `data.frame`s!



## Introducing the pipe (`%>%`)


Often times, we want to do multiple operations on our data and in a specific order.

For example I might want to do the following:


*Take* my dataset `smoke_complete` **and then**
*Sort* it by `cigarettes_per_day` **and then**
*filter* to have only `males` from the data.

The pipe (`%>%`) function acts like the **and then**:


```{r}
smoke_complete %>%
  arrange(cigarettes_per_day) %>%
  filter(gender == "male")

```


You can think of a pipe as putting the output of one step as an input of another. These two statements are equivalent:


```{r}
smoke_complete %>%
  arrange(cigarettes_per_day)
```

and

```{r}
arrange(smoke_complete, cigarettes_per_day)
```




## Sorting Data Frames using `arrange()`


Ok, we already have encountered `arrange()`. It's a function that lets us sort a `data.frame` by a variable.

By default, it sorts in *ascending* order:

```{r}
smoke_complete %>% 
  arrange(cigarettes_per_day)
```

```{r}
smoke_complete %>%
  arrange(desc(cigarettes_per_day))
```


You can also arrange by multiple variables. 

```{r}
smoke_complete %>%
  arrange(desc(cigarettes_per_day),
          tumor_stage)

```


Note that order of variables in `arrange()` matters!

```{r}
smoke_complete %>%
  arrange(
    tumor_stage,
    desc(cigarettes_per_day)
          )
```



## `filter()`ing our data


`filter()` is an extremely powerful function. It lets us subset our data according to specific criterion.

Let's filter on a `numeric` variable, `cigarettes_per_day`:


```{r}
smoke_complete %>%
  filter(cigarettes_per_day < 20)

```


We can also filter on a category. But that does require us to know the values of that categorical variable.


```{r}
smoke_complete %>%
  filter(tumor_stage == "stage iv")
```


## Filters a little logical

We can chain multiple criteria using the `&` (AND) or `|` (OR) operators. But we need to review a little logic before we do this.

If I wanted to return patients who were 

`male` **and** `stage iv`, 

I would want use an `&` to chain these criteria together:

```
gender == "male" & tumor_stage == "stage_iv"
```

```{r}
smoke_complete %>%
  filter(gender == "male" & 
           tumor_stage == "stage iv")

```


If I wanted patients who were

`male` **or** `stage iv`

I would use an `|` to chain these criteria together.

```{r}
smoke_complete %>%
  filter(gender == "male" |
           tumor_stage == "stage iv")
```


Think about it: which of the two code blocks will return a larger number of patients?



## Removing columns: using `select()`






## Errata

**If you would like more practice,** 
the [`exercises`](exercises/) directory contains additional tasks for practice.
Answers to those and in-class questions are available in the [`solutions`](solutions/) directory. 

**This document is written in [R markdown](http://rmarkdown.rstudio.com),**
which is a method of formatting text, code, and output to create documents that are sharable with other people. 
While this document is intended to serve as a reference for you to read while typing code into your own script,
you may also be interested in modifying and running code in the original R markdown file ([`class4.Rmd`](class4.Rmd) in the GitHub repository).
