---
title: "Assignment 3"
author: "You"
date: "12/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions for Submitting

Submit your knitted html file - if you are having issues, you can submit the `.Rmd` file, put please do your best to knit it. 

Make sure to name your assignment "assignment3-LASTNAME.html" and to fill out the space with you as an author above. 


# Assignment 3

1. Load the `smoke_complete` dataset from the `data` folder. Use `filter()` to subset those patients who have 

disease == "LUSC"

then save the results to an object called `lusc_smokers` (2 points):

```{r}
smoke_complete <- read_excel("data/smoke_complete.xlsx", sheet =1, na= "NA")


```

2. Sort `lusc_smokers` by decreasing `tumor_stage` (2 points):

```{r}


```


3. For `lusc_smokers`, make a boxplot where y = `cigarettes_per_day` and x = `tumor_stage`. (2 points)


```{r}


```

3. Modify your plot to facet by `gender`:


```{r}


```


4. This is a multi-part problem. (4 points)

Subset `smoke_complete` to a different set of patients, such as gender == "female".

Make a new faceted plot with `facet_wrap()`, this time a scatterplot. You're free to choose two numeric variables of interest to plot from `smoke_complete`. 

Color your plot by `disease` and facet by `vital_status`.


```{r}




```

5. Extra (2 bonus points): For one of the `geom` types in the BERD Workshop slides, produce a plot with `smoke_complete`.