---
title: "R Programming Part 1: Functions and objects"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vembedr)
```

# Before you get started

Please save a copy of this as `part1_FIRSTNAME_LASTNAME.Rmd` and work from that (where FIRSTNAME is your first name, and LASTNAME is your last name). This way, you'll have the original as a reference just in case.

Also, the first time you try something, try to *type* out the answer rather than cutting and pasting. It will help you understand what's going on, because it forces you to read the code.

# Welcome to R Programming!

This course introduces you to R in two parts:

Part 1 focuses on working through common tasks in data science:

-   importing,
-   wrangling,
-   visualizing, and
-   modeling data.

Part 2 focuses on more advanced topics:

-   automating analyses using metadata (`purrr`/for loops),
-   running basic statistical procedures,
-   machine learning using `tidymodels`,
-   and an introduction to Bioconductor for bioinformatics analysis.

Throughout, we'll work on concepts of reproducibility by utilizing project and notebook based-workflows and RMarkdown as a way of reproducibly sharing our work.

# Learning Objectives for the Course

Link to Video: <https://youtu.be/yCPcijCj15c?t=1502>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=1502")
```

# What is R?

R is a statistical and programming computer language widely used for a variety of applications.

Before proceeding with these training materials, please ensure you have an RStudio.cloud account and can see the workspace.

## Learning Objectives

By the end of this session, you should be able to:

1.  **Work** within the RStudio interface to **run** R code in an RMarkdown document
2.  **Understand** basic R syntax to **use** functions and **assign** values to objects
3.  **Create** and **manipulate** vectors and **understand** how R deals with missing data
4.  **Learn** about `for` loops as a way to process each value in a vector

## Introduction to R: Class 1

[Tour of R and R Notebooks]

Link to video: <https://youtu.be/yCPcijCj15c?t=3062>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=3062")
```

A good reference is here: `Help -> Cheatsheets -> RStudio IDE cheat sheet`

## Code Blocks

Link to video: <https://youtu.be/yCPcijCj15c?t=4019>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=4019")
```

Now that we have our notebook open, we're ready to start.

First thing of all, is the grey box below.

[Play button]

```{r}
# basic math
4 + 5 
```

Let's learn how to read this code block. Everything that starts with a `#` is called a *comment* and is not code that runs. It is useful for making notes for youself.

Below it is the actual code.

Try this one out. It's the same code as above, but with no spaces. Does it still run?

```{r}
# same code as above, without spaces
4+5
```

## Using Functions

Link to video: <https://youtu.be/yCPcijCj15c?t=4591>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=4591")
```

R includes functions for other types of math:

```{r}
# using a function: rounding numbers
round(3.14)
```

[function names and arguments, syntax]

```{r}
# using a function with more arguments
round(3.14, digits = 1)
```

[getting help]

```{r}
`?round`
```

[documentation and arguments]

```{r}
# can switch order of arguments
round(digits = 1, x = 3.14)
```

> You may notice that boxes pop up as you type. These represent RStudio's attempts to guess what you're typing and share additional options.
>
> You can also activate this by hitting the *Tab* key in the middle of typing. RStudio will attempt to complete, or fill in the rest of the characters.

### Challenge 1

what does the function `hist` do? What are its main arguments? How did you determine this?

```{r}

```

## Assigning objects

Link to video: <https://youtu.be/yCPcijCj15c?t=5118>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=5118")
```

```{r}
# assigning value to an object
weight_kg <- 55
```

[assignment operator, object naming conventions]

Now that the object has been assigned, we can reference that object by executing its name:

```{r}
# recall object
weight_kg
```

[using a variable]

```{r}
# multiple an object (convert kg to lb)
2.2 * weight_kg
```

[assigning a new variable]

```{r}
# assign weight conversion to object
weight_lb <- 2.2 * weight_kg
```

[environment panel]

```{r}
# reassign new value to an object
weight_kg <- 100
```

[changing variable values]

[objects that use other objects don't change]

> You can think of the names of objects like sticky notes. You have the option to place the sticky note (name) on any value you choose. You can pick up the sticky note and place it on another value, but you need to explicitly tell R when you want values assigned to certain objects.

```{r}
# remove object
remove(weight_lb) 
```

[removing objects]

> You can clear the entire environment using the button at the top of the Environment panel with a picture of a broom. This may seem extreme, but don't worry! We can re-create all the work we've already done by executing each line of code again.
>
> Recreating code from scratch is a big part of *analytic reproducibility*, in which you can give someone else your R code and data and they can recreate the analysis, include figures and tables.

### Challenge 2

What is the value of each item at each step? (Hint, you can see the value of an object by typing in the name of the object, such as with the `mass` line below.)

```{r}
mass <- 47.5            # 1. mass?
mass
width  <- 122             # 2. width?
mass <- mass * 2.0      # 3. mass?
width  <- width - 20        #4.  width?
mass_index <- mass/width  # 5. mass_index?
```

Make your answers here:

1.  

2.  

3.  

4.  

5.  

## Vectors

Link to video: <https://youtu.be/yCPcijCj15c?t=6049>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=6049")
```

[creating vectors] [c is for combine]

```{r}
# assign vector
ages <- c(50, 55, 60, 65) 
# recall vector
ages
```

[learning things about vectors]

```{r}
# how many things are in object?
length(ages)
```

```{r}
# what type of object?
class(ages)
```

```{r}
# performing functions with vectors
mean(ages)
range(ages)
```

[characters]

```{r}
# vector of body parts
organs <- c("lung", "prostate", "breast")
```

In this case, each word is encased in quotation marks, indicating these are character data, rather than object names.

### Challenge 3

Please answer the following questions about `organs`:

1.  How many values are in `organs`?
2.  What type of data is `organs`?
3.  get overview of `organs`

Answers here:

1.  

2.  

3.  

## Data types and Vectors

-   **character**: sometimes referred to as string data, tend to be surrounded by quotes
-   **numeric**: real or decimal numbers, sometimes referred to as "double"
-   **integer**: a subset of numeric in which numbers are stored as integers
-   **logical**: Boolean data (TRUE and FALSE)
-   **complex**: complex numbers with real and imaginary parts (e.g., 1 + 4i)
-   **raw**: bytes of data (machine readable, but not human readable)

### **Challenge 4:**

R tends to handle interpreting data types in the background of most operations.

The following code is designed to cause some unexpected results in R.

What is unusual about each of the following objects?

```{r}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```

## Manipulating vectors

Link to video: <https://youtu.be/yCPcijCj15c?t=6967>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=6967")
```

```{r}
# add a value to end of vector
ages <- c(ages, 90) 
```

```{r}
# add value at the beginning
ages <- c(30, ages)
```

```{r}
# extracting second value
organs[2] 
```

```{r}
# excluding second value
organs[-2] 
```

```{r}
# extracting first and third values
organs[c(1, 3)] 
```

## Missing data

```{r}
# create a vector with missing data
heights <- c(2, 4, 4, NA, 6)
```

[NA is not a character]

```{r}
# calculate mean and max on vector with missing data
mean(heights)
max(heights)
```

[ugh. na strikes again!]

```{r}
# add argument to remove NA
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

```{r}
# remove incomplete cases
na.omit(heights) 
```

**Challenge:** Complete the following tasks after creating this vector (Note: there are multiple solutions):

1.  Remove NAs on `more_heights` (assign it to the object `more_heights_complete`)
2.  Calculate the `median()` of `more_heights_complete`

```{r}
# create vector
more_heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
# remove NAs

# calculate the median

```

## Doing something with each element in a `vector`: `for` loops

Link to video: <https://youtu.be/yCPcijCj15c?t=7593>

```{r}
vembedr::embed_url("https://youtu.be/yCPcijCj15c?t=7593")
```

The last thing I want to cover today is how to do something with each element in a `vector`. If we do something once with a function, we can do it multiple times on different elements in a vector with that function. When we run the same function on different elements of a dataset, it is known as *iteration*.

Let's go through the code below.

The tricky thing to understand is that `height` is a *placeholder*. It changes each time we go through the loop, and it's a way to refer to the element of the list that time around.

```{r}
for(height in more_heights){
  print(height)
}
```

We won't be using `for` loops for a while, but I wanted you to see them so when you encounter them again, they're not a foreign concept.

## Wrapping up

[R/Rstudio] [objects] [data types]

## Post Class Survey

## Acknowledgements

This intro to R was adapted from material from Kate Hertweck's Intro to R course from <http://fredhutch.io>.
