---
title: "Answer Key Part 1"
author: "Ted Laderas"
date: "4/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Homework


### 1.  Object manipulation

Create an object called `agge` that contains your age in years. Then
reassign the object to a new object called `age` (e.g., correct the typo). Then remove the previous object from your environment and then calculate your age in days

```{r}
#create agge object
agge <- 43

#reassign agge value to age
age <- agge

#remove agge object
remove(agge)

#Use math to calculate your age in days
age * 365
```


### 2.  Vector manipulation (character data):

Create a object called `buildings` representing a vector that contains four names of buildings on OHSU's campus, including the building where you work (here's a reference: https://www.ohsu.edu/visit/maps). 

Add `Portland, Oregon` to the beginning of the vector, and `Phys Plant` to the end of the vector

subset the vector to show only the building in which you work.

```{r}
#create buildings object
buildings <- c("BICC", "Hatfield Research Center", "School of Nursing")
buildings
```

```{r}
#Add "Portland, Oregon" to the beginning of buildings
buildings <- c("Portland, Oregon", buildings)
buildings
```

```{r}
#Add "Phys Plant" to the end of buildings
buildings <- c(buildings, "Phys Plant")
buildings
```

```{r}
#subset buildings to show only your building
buildings[4]

```


## 3. Vector manipulation (numerical data):

The following vector represents the number of vacation days possessed by various employees.

How many employees are represented in the vector?

How many vacation days total?

```{r}
vacation_days <- c(5, 7, 20, 1, 0, 0, 12, 4, 2, 2, 2, 4, 5, 6, 7, 10, 4)

#how many employees are represented in the vector?
length(vacation_days)

#how many vacation days total?
sum(vacation_days)
```

