---
title: "Assignment 1"
author: "You"
date: "12/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment

### 1. Object manipulation

Create an object called `agge` that contains your age in years. Then reassign the object to a new object called `age` (e.g., correct the typo). Then remove the previous object from your environment and then calculate your age in days

```{r}
#create agge object

#reassign agge value to age

#remove agge object

#Use math to calculate your age in days

```

### 2. Vector manipulation (character data):

Create a object called `buildings` representing a vector that contains four names of buildings on OHSU's campus, including the building where you work (here's a reference: <https://www.ohsu.edu/visit/maps>).

Add `Portland, Oregon` to the beginning of the vector, and `Phys Plant` to the end of the vector

subset the vector to show only the building in which you work.

```{r}
#create buildings object
buildings <- c()

#Add "Portland, Oregon" to the beginning of buildings

#Add "Phys Plant" to the end of buildings

#subset buildings to show only your building
buildings[]

```

## 3. Vector manipulation (numerical data):

The following vector represents the number of vacation days possessed by various employees.

How many employees are represented in the vector?

How many vacation days total?

```{r}
vacation_days <- c(5, 7, 20, 1, 0, 0, 12, 4, 2, 2, 2, 4, 5, 6, 7, 10, 4)

#how many employees are represented in the vector?

#how many vacation days total?
```
