---
title: "More `case_when()` examples"
author: "Ted Laderas"
date: "2/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
data(penguins)
```

## More `case_when()` examples

In this document, we'll go over more `case_when()` examples.


## Making a continuous variable categorical: "short", "medium" and "long" billed penguins:

We want to be `exhaustive` in our coding. 

```{r}
coded_penguins <- penguins %>%
  mutate(bill_class = 
           case_when(
              bill_length_mm < 39 ~ "short",
              bill_length_mm >= 39 & bill_length_mm < 48 ~ "medium",
              bill_length_mm >= 48 ~ "long"
            )
         ) %>%
  select(species, island, bill_length_mm, bill_class)

coded_penguins
```

One way to check if you're exhaustive in your recoding is to plot your categorical variable on the x-axis and your numeric variable on the y-axis.

```{r}
ggplot(coded_penguins) +
  aes(x=bill_class, y=bill_length_mm) +
  geom_jitter()
```

## Combining categorical information

Note that both `species` and `island` are `factors`. Because of this, if we want to combine values from both of them, we have to cast them as `character()`.

```{r}
penguins %>%
  mutate(penguin_type = case_when(species == "Gentoo" ~ as.character(species),
                                  species == "Adelie" ~ as.character(island),
                                  TRUE ~ "other")) %>%
  select(species, island, penguin_type)

```