---
title: 'Part 5: Doing useful things with multiple tables/dates/functions/for-loops'
author: "you"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions for Submitting

Submit your knitted html file - if you are having issues, you can submit the `.Rmd` file, put please do your best to knit it. 

Make sure to name your assignment "assignment3-LASTNAME.html" and to fill out the space with you as an author above. 


## Assignment

1. Use `bind_rows()` to bind all three `smoke` datasets (`smoke_1`, `smoke_2`, and `smoke_3`) together. Hint: you'll have to do two `bind_row()` operations.

```{r}
smoke_1 <- read_excel(here('data/smoke_1.xlsx'), 
                      sheet=1, 
                      na="NA")
smoke_2 <- read_excel(here('data/smoke_2.xlsx'), 
                      sheet=1, 
                      na="NA")

smoke_3 <- read_excel(here('data/smoke_3.xlsx'), 
                      sheet=1, 
                      na="NA")
```



2. Take a look at the `patient` and `patient_encounter` tables loaded below.

```{r}
patient <- read_csv("data/patient_example.csv")
patient_encounter <- read_csv("data/patient_enc_example.csv")

```


How many rows does each table have?

```{r}
nrow()
nrow()
```


Write a `inner_join()` and assign it to `inner_patient` with 

- `patient` as the *left* table
- `patient_encounter` as the *right* table
- joining on the approriate columns (look at the column names carefully!)


```{r}
inner_patient <- patient %>%
  inner_join(y = ____, by = c())

inner_patient
```

How many rows does `inner_patient` have?


Do a `left_join` of the two tables with the same criteria as above. Assign it to `left_patient`. 

Count the number of rows in `left_patient`. Is it what you expected?

```{r}
left_patient <- patient %>%
  left_join()

nrow(left_patient)
```


Look up the documentation for `anti_join()`. Do an `anti_join` of the two tables with the same criteria. What does this joined table contain?

```{r}
anti_patient <- patient %>%
  anti_join()

nrow(anti_patient)
```


3. Look up the documentation for `tidyr::pivot_wider`. Transform `expression_long` back to a *wide* format, but this time with **columns** as *genes* and **rows** as *samples*.

Hint: Always look at the examples in the documentation! They'll help you.

```{r}
expression_example <- read_excel("data/expression_example.xlsx", sheet=1)

expression_long <- 
  expression_example %>%
    pivot_longer(cols= starts_with("tcga"), 
                 names_to = "sample", 
                 values_to = "expression")



expression_long %>%
  pivot_wider(id_cols = ------,
              names_from = ------,
              values_from = ------) 


```

4. What does the function `load_files()` do? 

Write up a short explanation here:



```{r}
load_files <- function(path){
  out_frame <- readxl::read_excel(path, sheet = 1)
  out_frame <- janitor::clean_names(out_frame)
  return(out_frame)
}
```

Write up an example here (You can use one of the excel files in the `data` folder).

```{r}

```

5. Write a function called `count_columns()` that takes a `data.frame` as input and returns the number of columns in the `data.frame`

```{r}
count_columns <- function(df){
  
  
}
```

Show an example of your function in use:

```{r}


```



## Acknowledgements

Written by Aaron Coyner and Ted Laderas. Data is derived from Intro to R from fredhutch.io. 
