---
title: 'Part 4: mutate(), group_by/summarize()'
author: "you"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives

- **Learn** and apply `mutate()` to change the data type of a variable
- **Apply** `mutate()` to calculate a new variable based on other variables in a `data.frame`.
- **Apply** `case_when` in a `mutate()` statement to make a continuous variable categorical
- **Apply** `group_by()/summarize()` as a pattern to get summary statistics, including counts, means, and standard deviations within a category.
- **Standardize** variable names using `clean_names()`


## Getting set up

```{r prep, message=FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(readxl)
library(janitor)


smoke_complete <- read_excel("data/smoke_complete.xlsx", 
                             sheet=1, 
                             na="NA")
```



# `mutate()` - A confusing name, a powerful `dplyr` verb


So what is mutate? `mutate()` is one of the most useful `dplyr` verbs. You can use it to transform data (variables in your `data.frame`) and add it as a new variable into the data.frame.

I tend to think of this much like adding a formula in Excel to calculate the value of a new column based on previous columns. You can do lots of things such as:

- subtract one column from another
- convert the units of one column to new units (such as days to years)
- change the capitalization of categories in a variable
- recode a continuous variable to be a categorical one




## Using `mutate` to calculate a new variable based on other variables

One use for `mutate` is to do Excel type calculations using other columns on the data. For instance, we might want to calculate the sum of `age_at_diagnoses`and `days_to_death` to get the `age_at_death`.

```{r}
smoke_complete %>% 
    mutate(age_at_death = age_at_diagnosis + days_to_death) %>%
    glimpse()
```


## Your Turn

Create a variable called `cigarettes_total` by multiplying `cigarettes_per_day` by `-days_to_birth` below:

```{r}
smoke_complete %>%
    mutate(cigarettes_total = ) 

```




## Using `mutate` to make our `character` variables into `factors`

Remember `factor`s? Well, we can use `mutate()` to make a `character` variable a `factor`.

Let's convert `gender` from `character` into `factor`. We'll then pipe the output into `glimpse()` so you can see the variable types.



```{r}
smoke_complete %>% 
    #reassign the gender variable to be a factor
    mutate(gender = factor(gender)) %>%
    glimpse()
```


One thing to notice: we are doing something called *reassignment* here. We're taking the previous values of our variable (`gender`), doing something to it (making it a `factor`), and then reassigning the variable `gender` to our fixed set of values.


Remember when we wanted to decide the order of the categories in the factor? We can assign that order using the `levels` argument in `factor()`. To show the order, we can pipe the output into a command called `tabyl()`, from the `janitor` package, which is a `tidyverse` friendly version of `table()`.


```{r}
smoke_complete %>% 
    #reassign the gender variable to be a factor
    mutate(gender = factor(gender, levels = c("female", "male"))) %>%
    tabyl(gender)
```

Notice that the `female` value goes before the `male`, which is what we wanted.


## Using `mutate` to replace missing values with `replace_na`

Sometimes we want to fill in missing values with a certain value. We can use the `replace_na()` function inside of `mutate()` to specify this. For example, if the days to last follow up is missing, we want to set their days to follow up to 0. The first two observations in this data have missing follow up, but when we replace them we note they are now equal to 0:

```{r}
smoke_complete %>%
    mutate(days_to_last_follow_up = replace_na(days_to_last_follow_up, 0)) %>%
    glimpse()
```



## Using `mutate` to make a continuous variable categorical using `case_when`


Say we want to make the `cigarettes_per_day` into a categorical variable with the values:


* 0-5 cigarettes/day
* 6+ cigarettes/day


How would we do that?

Well, we need to follow the basic pattern for each of our categories:

*condition* ~ *category name*


The *left* side of the `~` is where we can specify how we define the *category&. 

The *right* side of the `~` is where we can specify the *category name* (as a character).

In the example below,

`cigarettes_per_day <= 5` is our left side,
and `0-5` is our right side (our category).

We need to do this for each level in our category.

```{r}
smoke_complete %>% 
    mutate(
        cigarettes_category = case_when(
            cigarettes_per_day <= 5 ~ '0-5',
            cigarettes_per_day > 5 ~ '6+'
            )
        ) %>%
    mutate(cigarettes_category = 
               factor(cigarettes_category,
                      levels = c("0-5", "6+")
                      )
           ) %>%
    tabyl(cigarettes_category)
    
```



## Your Turn

Modify the code below to recode `cigarettes_category` to have 3 levels:

* `0-5` 
* `6-10` 
* `11+`

Hint: you'll have to chain conditions with a `&` to get the `6-10` category.


```{r}
smoke_complete %>% 
    mutate(
        cigarettes_category = case_when(
            cigarettes_per_day <= 5 ~ '0-5',
            cigarettes_per_day > 5 ~ '6+'
            )
        ) %>%
    mutate(cigarettes_category = 
               factor(cigarettes_category,
                      levels = c("0-5", "6+")
                      )
           ) %>%
    tabyl(cigarettes_category)
    
```





# `summarize()`

The `summarize()` verb produces summaries across your data. 

For example, if I wanted to find the average `cigarettes_per_day` across all my data, I'd do this:


```{r}
smoke_complete %>%
    summarize(average_cigarettes = mean(cigarettes_per_day))

```

Note we need to be careful if there are missing values in the variable we are summarizing, compare these two results:

```{r}
smoke_complete %>%
    summarize(average_yrs_smoked = mean(years_smoked))

smoke_complete %>%
    summarize(average_yrs_smoked = mean(years_smoked, na.rm = TRUE))

```

What are the kinds of things that summarize is useful for?

Useful functions (from https://dplyr.tidyverse.org/reference/summarise.html)

Center: `mean()`, `median()`
Spread: `sd()` (standard deviation) 
Range: `min()`, `max()` 
Position: `first()`, `last()`, `nth()`
Count: `n()`, `n_distinct()`



By itself, `summarize()` is ok. But it's really combined with `group_by()` that it becomes extremely powerful.



# `group_by()/summarize()`


These two verbs always go together. `group_by()` doesn't do anything by itself, output wise. But what it does is split our data frame into a number of smaller `data.frame`s split out by category. Then we use `summarize` to do some sort of summary calculation (such as counting or calculating the mean) on these smaller `data.frame`s:


For example, if we want to calculate the mean `cigarettes_per_day` smoked between `male`s and `female`s, we'd do the following:

```{r}
smoke_complete %>%
   group_by(gender) %>%
   summarize(mean_cigarettes = mean(cigarettes_per_day))
```




## `group_by()/summarize` to calculate mean and standard deviation values

Below, we'll calculate the average number of cigarettes smoked per day for males and females, and the standard deviation.


```{r}
smoke_complete %>%
    group_by(gender) %>%
    summarize(mean = mean(cigarettes_per_day), 
              sd = sd(cigarettes_per_day))
```




## `group_by()` to `count()` things

What if we want to know the number of subjects in our dataset with each type of `disease`? We can use the `n()` function.

```{r}
smoke_complete %>%
    group_by(disease) %>%
    summarize(count = n())
```

We can also use the `count()` function directly:

```{r}
smoke_complete %>%
    count(disease)
```



## Your Turn

Use `group_by()/summarize()` to calculate the maximum cigarettes per day within each `disease` category.


```{r}
smoke_complete %>%
    group_by(------) %>%
    summarize(max_cigarettes = -------)

```


# `summarize()` and `mutate()` across multiple columns with `across()`

Often we may want to summarize or mutate multiple variables in the same way. We can follow the examples above and specify each operation individually. For instance, suppose we want to calculate the mean of multiple numeric variables:

```{r}
smoke_complete %>%
    summarize(mean_days_to_death = mean(days_to_death, na.rm = TRUE),
              mean_days_to_birth = mean(days_to_birth),
              mean_days_to_last_followup = mean(days_to_last_follow_up, na.rm = TRUE))
```
Phew, that's a lot of typing! But there's an easier way:

```{r}
smoke_complete %>%
    summarize(
        across(c(days_to_death, days_to_birth, days_to_last_follow_up), 
               mean, na.rm = TRUE))
```
The `across()` function used above, first asks which variables, then asks which operation. In the example above, we named the variables first, and then we named the function `mean` and added an argument `na.rm = TRUE` which is applied to the mean function. If we had not done this, we would have `NA`s in our results.

We can also use the syntax from `select()` to choose our variables, for instance, we could call the same variables by using `starts_with("days")`:

```{r}
smoke_complete %>%
    summarize(
        across(starts_with("days"), 
               mean, na.rm = TRUE))
```

Or we can calculate the mean on all numeric variables:

```{r}
smoke_complete %>%
    summarize(
        across(where(is.numeric), 
               mean, na.rm = TRUE))
```

Using `across()` takes some getting used to, but seeing more examples will help you see how it works. There are also more complex ways to use your own functions and naming conventions. Look at this vignette for many examples:  [dplyr colwise vignette](https://dplyr.tidyverse.org/articles/colwise.html).


For an example of using our own function, we calculate the number of `NA`s across all columns with `everything()` and a custom function defined with `~` operating on the argument `.x`:

```{r}
smoke_complete %>%
    summarize(across(everything(), ~sum(is.na(.x)))) %>%
    glimpse()
```

We can also use `across()` with `mutate()` in a similar way. For example, if we want to convert all character columns to factors:

```{r}
smoke_complete %>%
    mutate(
        across(where(is.character), factor)) %>%
    glimpse()
```

## Your turn

Convert all variables that start with `days_to` into years:

```{r}
smoke_complete %>%
    mutate(
        across(starts_with(----), ~ .x/-----)) %>%
    glimpse()
```

# Standardizing variable names: `clean_names()`

The last package we'll talk about today is the `janitor` package. This package adds some really useful functions for *cleaning* data.

The one I use the most is called `clean_names()`. This will mutate column names into a standard format - all character values will be lowercase and spaces will be converted to underscores. 

The `smoke_complete` dataset has already had this function applied to it, but let's create a smaller example dataset where this wouldn't be the case.


```{r}
smoke_bad_colnames <- read_csv("data/smoke_bad_colnames.csv")
smoke_bad_colnames
```

Ack! These column names have spaces, inconsistent capitalization, which make these difficult to work with.

So let's use `clean_names()` to fix this:


```{r}
smoke_good_colnames <- smoke_bad_colnames %>%
    clean_names()

smoke_good_colnames
```

By default, `clean_names()` converts to snake case (spaces between words are replaced by an underscore `_`).  `clean_names()` will help you standardize your dataset's column names.

## Using `rename()` to rename!

We can easily change names by hand, as well, with the `rename()` function in `dplyr` which is essentially a wrapper for `mutate()`:

```{r}
smoke_good_colnames %>%
    rename(age = age_at_diagnosis) %>%
    glimpse()
```


## Assignment 4

1. Load the `smoke_complete` dataset from the `data` folder.

```{r}

    
```


2. Determine which tissue/organ of origin occurs most frequently in `smoke_complete`. (Hint: use both `group_by()/summarize()` and `arrange()`.)

```{r}
smoke_complete %>%
    
```


3.  Create a new dataset called `smoke_filtered` that only contains subjects with `tissue_or_organ_of_origin` of `C34.3`. Use `glimpse()` on `smoke_filtered`.

```{r}
smoke_filtered  <- smoke_complete %>%
    

glimpse(smoke_filtered)
```

4. For subjects in `smoke_filtered`, make a new variable called `age_years` by converting `age_at_diagnosis` from days to years.

```{r}
smoke_filtered <- smoke_filtered %>% 
    mutate(age_years = _______ * _____)

```


4. Make a new variable called `age_category` using `case_when()` from `age_years` in `smoke_filtered`. 

Make `age_category` a `factor` with the following levels:

* `0-40` years old
* `40-55` years old
* `55-70` years old
* `70+` years old


```{r}
smoke_filtered <- 
    smoke_filtered %>% 
    mutate(age_category = 
               case_when(age_years <= 40 ~ "0-40",
                         
                         
                         
                         )
           ) %>%
    mutate(age_category = factor(age_category,
                                 levels = c("0-40",
                                            
                                            
                                            
                                            )
                                 )
           )


```


5. Create a scatterplot using `geom_point()` demonstrating the correlation (if any) between the categorical variable `age_at_diagnosis` versus `cigarettes_per_day`.


```{r}
ggplot(smoke_complete) +
    aes(x = , y= ) +
    geom_point()

```


6. Create a column called `survival_time` that is equal to `days_to_death` if non=missing, and is equal to `days_to_last_follow_up` if missing.

```{r}
smoke_filtered <- 
    smoke_filtered %>% 
    mutate(survival_time = )
```

7. Create a column called `survival_event` that is 1 if the person has a non missing `days_to_death` and 0 if not.

```{r}
smoke_filtered <- 
    smoke_filtered %>% 
    mutate(survival_event = )
```

8. Calculate median `survival_time` in patients who died versus those who were lost to follow up. Be careful with missing values in `survival_time`.

```{r}
smoke_filtered %>% 
    group_by(   ) %>%
    summarize(median_surv_time = )
```

9. Calculate median `survival_time` in patients who died versus those who were lost to follow up, stratified by disease type. Hint: `group_by()` can take multiple arguments to group by multiple variables.

```{r}
smoke_filtered %>% 
    group_by(  , ) %>%
    summarize(median_surv_time = )
```

10. Create a boxplot of `survival_time` by `factor(survival_event)` with color mapped to `disease`.

```{r}
ggplot(smoke_filtered, 
       aes(x = factor( ), 
           y =  , 
           color =  )) + 
    geom_boxplot()
```


# Acknowledgments

Written by Aaron Coyner and Ted Laderas and Jessica Minnier.

Based on the Intro to R materials from fredhutch.io and the R-Bootcamp
