---
title: "Assignment 4"
author: "You"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Assignment 4

1. Load the `smoke_complete` dataset from the `data` folder.

```{r}

    
```


2. Determine which tissue/organ of origin occurs most frequently in `smoke_complete`. (Hint: use both `group_by()/summarize()` and `arrange()`.)

```{r}
smoke_complete %>%
    
```


3.  Create a new dataset called `smoke_filtered` that only contains subjects with `tissue_or_organ_of_origin` of `C34.3` or `C53.9`. Use `glimpse()` on `smoke_filtered`.

Hint: you can use `filter()` with the `%in%` operator like this:

```
penguins %>%
   filter(species %in% c("Adelie", "Chinstrap"))

```


```{r}
smoke_filtered  <- smoke_complete %>%
    

glimpse(smoke_filtered)
```

4. For subjects in `smoke_filtered`, make a new variable called `age_years` by converting `age_at_diagnosis` from days to years.

```{r}
smoke_filtered <- smoke_filtered %>% 
    mutate(age_years = _______ * _____)

```


4. Make a new variable called `age_category` using `case_when()` from `age_years` in `smoke_filtered`. 

Make `age_category` a `factor` with the following levels:

* `0-40` years old
* `40-55` years old
* `55-70` years old
* `70+` years old


```{r}
smoke_filtered <- 
    smoke_filtered %>% 
    mutate(age_category = 
               case_when(age_years <= 40 ~ "0-40",
                         
                         
                         
                         )
           ) %>%
    mutate(age_category = factor(age_category,
                                 levels = c("0-40",
                                            
                                            
                                            
                                            )
                                 )
           )


```


5. Create a scatterplot using `geom_point()` demonstrating the correlation (if any) between the categorical variable `age_at_diagnosis` versus `cigarettes_per_day`.


```{r}
ggplot(smoke_complete) +
    aes(x = , y= ) +
    geom_point()

```


6. Create a column called `survival_time` that is equal to `days_to_death` if non=missing, and is equal to `days_to_last_follow_up` if missing.

Hint: you can use `is.na()` in your `case_when()` statement. 

If you want non-`NA` values, you can use `!is.na()` (the `!` is short for Not, so this means not-NA values)


```{r}
smoke_filtered <- 
    smoke_filtered %>% 
    mutate(survival_time = 
    #           
               )
```

7. Create a column called `survival_event` that is 1 if the person has a non missing `days_to_death` and 0 if not.

```{r}
smoke_filtered <- 
    smoke_filtered %>% 
    mutate(survival_event = )
```

8. Calculate median `survival_time` in patients who died versus those who were lost to follow up (use `survival_event` in your `group_by()`). Be careful with missing values in `survival_time`.

```{r}
smoke_filtered %>% 
    group_by(   ) %>%
    summarize(median_surv_time = )
```

9. Calculate median `survival_time` in patients who died versus those who were lost to follow up (again, use `survival_event`), stratified by disease type. Hint: `group_by()` can take multiple arguments to group by multiple variables.

```{r}
smoke_filtered %>% 
    group_by(  , ) %>%
    summarize(median_surv_time = )
```

10. Create a boxplot of `survival_time` by `factor(survival_event)` with `fill` mapped to `disease`.

```{r}
ggplot(smoke_filtered, 
       aes(x = factor( ), 
           y =  , 
           fill =  )) + 
    geom_boxplot()
```
