---
title: 'Part 4: mutate, group_by/summarize'
author: "you"
date: "5/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives

- **Learn** and apply `mutate()` to change the data type of a variable
- **Apply** `mutate()` to calculate a new variable based on other variables in a `data.frame`.
- **Apply** `case_when` in a `mutate()` statement to make a continuous variable categorical
- **Apply** `group_by()/summarize()` as a pattern to get summary statistics, including counts, means, and standard deviations within a category.
- **Identify** and **transform** wide form data to long form data using `pivot_longer()`
- **Standardize** variable names using `clean_names()`



# `mutate()` - A confusing name, a powerful `dplyr` verb


So what is mutate? `mutate()` is one of the most useful `dplyr` verbs. You can use it to transform data (variables in your `data.frame`) and add it as a new variable into the data.frame.

I tend to think of this much like adding a formula in excel to calculate the value of a new variable 




## Using `mutate` to make our `character` variables into `factors`

Remember `factor`s? Well, we can use `mutate()` to make a `character` variable a `factor`.


```{r}

```



## Using `mutate` to calculate a new variable based on other variables

Another use for `mutate` is to do Excel type calculations using other columns on the data. For instance, we might want to calcluate the difference between `age_at_diagnoses`and 

```{r}

```




## Using `mutate` to make a continuous variable categorical using `case_when`

Say we want to make the `cigarettes_per_day` into a categorical variable with the values


* 0-1 cigarettes/day
* 2-5 cigarettes/day
* 6 or more cigarettes/day


How would we do that? 


```{r}

```



# `group_by()/summarize()`

These two verbs always go together.



## `group_by()/summarize` to calculate mean and standard deviation values

```{r}

```




## `group_by()/summarize()` to count things


```{r}


```


# Wide versus long data: `pivot_longer()`




# Standardizing variable names: `clean_names()`

The last package we'll talk about today is the `janitor` package. This package adds some really useful functions for *cleaning* data.

The one I use the most is called `clean_names`. 


```{r}

```


```{r}


```




